---
title: "Data File"
author: "Mark Nathin and Anqi Liu"
date: "November 14, 2021"
output: 
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
---

```{r, setup, include=FALSE}
require(mosaic)   # Load additional packages here 
knitr::opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
```

```{r}
library(readr)
library(dplyr)
library(readxl)
```


"I think the motivation for using web data could be highlighted a bit more - eventually, all data on slide 4 could also come from surveys. So for the GTrends data, would the argument be that this is accessible over time (as opposed to a single point in time). If yes, maybe this aspect could be incorporated in your models (e.g. change in mask wearing search interest up to the election)."
  


```{r}
library(tidycensus)
acs5 <- load_variables(year = 2017, dataset = "acs5", cache = TRUE)
```

```{r}
race <- get_acs(geography = "county",
variables = c(RaceTotal = "B02001_001",
White = "B02001_002",
Black = "B02001_003",
AmericanIndianAlaskaNative = "B02001_004",
Asian = "B02001_005",
NativeHawaiian = "B02001_006",
OtherRace = "B02001_007",
TwoRaces = "B02001_008"),
survey = "acs5")
```


## Mask-Usage Data 

First we will read-in the NY Times Mask-Usage data from July 2020. 

```{r}
mask_use_by_county <- read_csv("~/Desktop/mask-use-by-county.csv")
mask_use <- mask_use_by_county
head(mask_use)
```

## Election Results Data 

Next we will read in the county level results of the 2020 Presidential Election. 

```{r}
X2020_US_County_Level_Presidential_Results <- read_csv("~/Desktop/2020_US_County_Level_Presidential_Results.csv")
Election <- X2020_US_County_Level_Presidential_Results
Election <- rename(Election, COUNTYFP = county_fips)
head(Election)
```

## Income and other County Data 

```{r}
library(readxl)
Unemployment <- read_excel("~/Desktop/Unemployment.xlsx")
```

```{r}
county_data <- read_csv("~/Downloads/county_complete.csv")
```

```{r}
Education <- read_excel("~/Desktop/Education.xls")
```


## Joining Data

Mask usage data was not collected for any counties in Alaska. Therefore, Alaska will not be included in this analysis. 

### Mask and Election Data Joining

```{r}
Left <- left_join(Election, mask_use, by = "COUNTYFP")
Combined <- inner_join(Election, mask_use, by = "COUNTYFP")
```


## Preliminary Analysis

Which counties in the United States were most and least likely to report wearing a mask in public during the height of the pandemic?

```{r}
Combined %>% # Most likely to always wear mask
  select(state_name, county_name, ALWAYS) %>%
  group_by(state_name, county_name, ALWAYS) %>%
  summarize(High_Masking = n()) %>%
  arrange(desc(ALWAYS))

Combined %>% # Most likely to never wear mask
  select(state_name, county_name, NEVER) %>%
  group_by(state_name, county_name, NEVER) %>%
  summarize(Low_Masking = n()) %>%
  arrange(desc(NEVER))
```

--> Could create new variables: Combining Never/Rarely and Freuently/Always into one variable.


What is the link between mask usage and 2020 Presidential election votes: how well can one predict the other?  

### Predict Mask-Wearing with Percentage Vote for Democrats/Republican

```{r}
mod1 <- lm(ALWAYS ~ per_dem + per_gop, data = Combined)
summary(mod1)
faraway::vif(mod1) # Correlated predictors - take one out
```

```{r}
mod2 <- lm(NEVER ~ per_dem + per_gop, data = Combined)
summary(mod2)
```

### Predict Percentage Vote for Democrats/Republican with Mask Wearing Tendencies

```{r}
mod3 <- lm(per_dem ~ ALWAYS + NEVER, data = Combined)
summary(mod3)
```

```{r}
mod4 <- lm(per_gop ~ ALWAYS + NEVER, data = Combined)
summary(mod4)
faraway::vif(mod4) # Predictors not correlated 
```

Always wearing a mask is highly associated with a high vote percentage for democratic candidates. Never wearing a mask is associated with voting republican but the association is less strong (hypothesis: perhaps because a lot of people were wearing masks overall- perhaps fewer Republicans reported wearing a mask than Democrats but the majority of Republicans still wore masks)

How much can other variables improve these types of models? 

## Google Trends

```{r}

```


--> Use Census API for acquiring additional county level data? 

## References

https://www.ers.usda.gov/data-products/county-level-data-sets/download-data

Mask-wearing data: https://github.com/nytimes/covid-19-data/tree/master/mask-use

Election data: https://github.com/tonmcg/US_County_Level_Election_Results_08-20/blob/master/2020_US_County_Level_Presidential_Results.csv







