---
title: "Data File"
author: "Mark Nathin and Anqi Liu"
date: "November 14, 2021"
output: 
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
---

```{r, setup, include=FALSE}
require(mosaic)   # Load additional packages here 
knitr::opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
```

```{r}
library(readr)
library(dplyr)
library(readxl)
```

## Mask-Usage Data 

First we will read-in the NY Times Mask-Usage data from July 2020. 

```{r}
mask_use_by_county <- read_csv("~/Desktop/mask-use-by-county.csv")
mask_use <- mask_use_by_county
head(mask_use)
```

## Election Results Data 

Next we will read in the county level results of the 2020 Presidential Election. 

```{r}
X2020_US_County_Level_Presidential_Results <- read_csv("~/Desktop/2020_US_County_Level_Presidential_Results.csv")
Election <- X2020_US_County_Level_Presidential_Results
Election <- rename(Election, COUNTYFP = county_fips)
head(Election)
```

## Census API Variables

```{r}
library(tidycensus)
acs5 <- load_variables(year = 2017, dataset = "acs5", cache = TRUE)
```

```{r}
race <- get_acs(geography = "county",
variables = c(RaceTotal = "B02001_001",
White = "B02001_002",
Black = "B02001_003",
AmericanIndianAlaskaNative = "B02001_004",
Asian = "B02001_005",
NativeHawaiian = "B02001_006",
OtherRace = "B02001_007",
TwoRaces = "B02001_008"),
survey = "acs5")
```

```{r}
spRace <- race %>% select(-moe) %>% spread(variable, estimate)
```

```{r}
# Standardizing variables by county population
RaceCleaned <- spRace %>% mutate(White = White/RaceTotal,
Black = Black/RaceTotal,
AmericanIndianAlaskaNative = AmericanIndianAlaskaNative/RaceTotal,
Asian = Asian/RaceTotal,
NativeHawaiian = NativeHawaiian/RaceTotal,
TwoRaces = TwoRaces/RaceTotal) %>%
select(GEOID, NAME, RaceTotal, White, Black, AmericanIndianAlaskaNative,
Asian, NativeHawaiian, TwoRaces)
```

```{r}
EmploymentIncome <- get_acs(geography = "county",
                      variables = c(OverAge16 = "B23025_001",
                                    UnemploymentRate = "B23025_005",
                                    Med_Household_Income = "B19019_001",
                                    PerCapitaIncome = "B19301_001"),
                      survey = "acs5")
```

```{r}
spEmploymentIncome <- EmploymentIncome %>% select(-moe) %>% spread(variable, estimate)
```

```{r}
# Standardizing variables by county population
EmploymentIncomeCleaned <- spEmploymentIncome %>% mutate(UnemploymentRate = UnemploymentRate/OverAge16) %>%
select(GEOID, NAME, UnemploymentRate, Med_Household_Income, PerCapitaIncome)
```

```{r}
education <- get_acs(geography = "county",
                     variables = c(TotOver25 = "B07009_001", 
                                   LessHS = "B07009_002", 
                                   HSGrad = "B07009_003",
                                   SomeCollege = "B07009_004", 
                                   Bachelors = "B07009_005", 
                                   GradProfDegree = "B07009_006"), survey = "acs5")
```

```{r}
spEducation <- education %>% select(-moe) %>% spread(variable, estimate)
```

```{r}
EducationCleaned <- spEducation %>% mutate(LessHS = LessHS/TotOver25,
HSGrad = HSGrad/TotOver25,
SomeCollege = SomeCollege/TotOver25,
Bachelors = Bachelors/TotOver25,
GradProfDegree = GradProfDegree/TotOver25) %>%
  select(GEOID, NAME, HSGrad, SomeCollege, Bachelors, GradProfDegree)
```

```{r}
Join1 <- inner_join(RaceCleaned, EmploymentIncomeCleaned, by = "NAME")
Join2 <- inner_join(Join1, EducationCleaned, by = "NAME")
Join2 <- Join2 %>%
  select(-GEOID.y) %>%
  rename(COUNTYFP = "GEOID.x")
```


## Joining Data

Mask usage data was not collected for any counties in Alaska. Therefore, Alaska will not be included in this analysis. 

### Mask and Election Data Joining

```{r}
Left <- left_join(Election, mask_use, by = "COUNTYFP")
Combined <- inner_join(Election, mask_use, by = "COUNTYFP")
```

```{r}
uscounties <- read_csv("~/Downloads/simplemaps_uscounties_basicv1.71/uscounties.csv")
uscounties <- uscounties %>%
  rename(COUNTYFP = county_fips)
Combined2 <- inner_join(Combined, uscounties, by = "COUNTYFP")
```

```{r}
CombinedClean <- Combined2 %>%
    select(state_name.x, COUNTYFP, county_name, per_gop, per_dem, per_point_diff, NEVER, RARELY, SOMETIMES, 
           FREQUENTLY, ALWAYS, lat, lng, population) %>%
  rename(Percent_Democrat = "per_dem") %>%
  rename(Percent_Republican = "per_gop") %>%
  rename(Difference = "per_point_diff") %>%
  rename(State = "state_name.x") %>%
  rename(Never = "NEVER") %>%
  rename(Rarely = "RARELY") %>%
  rename(Sometimes = "SOMETIMES") %>%
  rename(Frequently = "FREQUENTLY") %>%
  rename(Always = "ALWAYS") %>%
  mutate(Never_Rarely = Never + Rarely) %>%
  mutate(Frequently_Always = Frequently + Always) %>%
  mutate(Majority = ifelse(Percent_Democrat > Percent_Republican, "Democrat", "Republican"))

CombinedClean$Percent_Democrat <- round(CombinedClean$Percent_Democrat, digits = 2)
CombinedClean$Percent_Republican <- round(CombinedClean$Percent_Republican, digits = 2)
CombinedClean$Difference <- round(CombinedClean$Difference, digits = 2)
```

```{r}
CombinedClean <- left_join(CombinedClean, Join2, by = "COUNTYFP")
```


--> ROUND TO TWO DIGITS 
```{r}
#CombinedClean$Percent_Democrat <- round(CombinedClean$Percent_Democrat, digits = 2)
#CombinedClean$Percent_Republican <- round(CombinedClean$Percent_Republican, digits = 2)
#CombinedClean$Difference <- round(CombinedClean$Difference, digits = 2)
```


```{r}
ShinyAppDS <- CombinedClean %>% 
  select(State, county_name, lat, lng, Percent_Republican, Percent_Democrat, Difference, Never, Rarely, 
         Sometimes, Frequently, Always, Never_Rarely, Frequently_Always, Majority)
numcounties <- 700
ShinyAppDS <- ShinyAppDS[sample(1:nrow(ShinyAppDS), numcounties, replace=FALSE),] # Select a random sample of 700 counties
ShinyAppDS

write_csv(ShinyAppDS, "ShinyAppDS.csv")
```

## Preliminary Analysis

Which counties in the United States were most and least likely to report wearing a mask in public during the height of the pandemic?

```{r}
Combined %>% # Most likely to always wear mask
  select(state_name, county_name, ALWAYS) %>%
  group_by(state_name, county_name, ALWAYS) %>%
  summarize(High_Masking = n()) %>%
  arrange(desc(ALWAYS))

Combined %>% # Most likely to never wear mask
  select(state_name, county_name, NEVER) %>%
  group_by(state_name, county_name, NEVER) %>%
  summarize(Low_Masking = n()) %>%
  arrange(desc(NEVER))
```


What is the link between mask usage and 2020 Presidential election votes: how well can one predict the other?  

```{r}
cor(CombinedClean$Never_Rarely, CombinedClean$Percent_Republican)
cor(CombinedClean$Frequently_Always, CombinedClean$Percent_Democrat)
cor(CombinedClean$Frequently_Always, CombinedClean$PerCapitaIncome)
cor(CombinedClean$Frequently_Always, CombinedClean$Percent_Democrat)
```


### Predict Mask-Wearing with Percentage Vote for Democrats/Republican

```{r}
# Simple Linear Regression
NoMask <- lm(Never_Rarely ~ Percent_Republican, data = CombinedClean)
summary(NoMask)
```

```{r}
# Simple Linear Regression
YesMask <- lm(Frequently_Always ~ Percent_Democrat, data = CombinedClean)
summary(YesMask)
```



```{r}
MLRMod <- lm(Frequently_Always ~ Percent_Democrat + White + Asian + Black + PerCapitaIncome + Bachelors, data = CombinedClean)
summary(MLRMod)
```


### Predict Percentage Vote for Democrats/Republican with Mask Wearing Tendencies

```{r}
## Simple Linear Regression
DemocratSupport <- lm(Percent_Democrat ~ Frequently_Always, data = CombinedClean)
summary(DemocratSupport)
```

```{r}
# Simple Linear Regression
RepubSupport <- lm(Percent_Republican ~ Never_Rarely, data = CombinedClean)
summary(RepubSupport)
```

Always wearing a mask is highly associated with a high vote percentage for democratic candidates. Never wearing a mask is associated with voting republican but the association is less strong (hypothesis: perhaps because a lot of people were wearing masks overall- perhaps fewer Republicans reported wearing a mask than Democrats but the majority of Republicans still wore masks)


```{r}
Mult <- lm(Percent_Democrat ~ Frequently_Always + White + Asian + Black + PerCapitaIncome + Bachelors, 
           data = CombinedClean)
summary(Mult)
```








## References

https://www.ers.usda.gov/data-products/county-level-data-sets/download-data

Mask-wearing data: https://github.com/nytimes/covid-19-data/tree/master/mask-use

Election data: https://github.com/tonmcg/US_County_Level_Election_Results_08-20/blob/master/2020_US_County_Level_Presidential_Results.csv







